---
title: "Hibernationslowsepigeneticaging"
author: "Isabel Sullivan"
date: "10/20/2021"
output: html_document
---

```{r}
library(data.table)
library(readr)
library(readxl)
library(tidyverse)
library(dplyr)
library(glmnet)
library(Matrix)
library(broom)
library(caret)
library(lattice)
set.seed(123)

```

This code performs a comparative analysis of winter to summer DNA samples from the same Eptesicus fuscus individuals. 

To control for age effects, start by taking the residuals from the age relationship (i.e.regression) at each site. Using ages estimated by the epigenetic clock (Wilkinson et al., 2021), modeled (methylation value ~ Age) grouped per site. Then, take the residuals from this regression at each site. For each analysis, the residuals for bats used in said analysis are pulled.

```{r}
get(load("all_probes_sesame_normalized.Rdata")) #this has beta values for each bat at each CpG site
raw<-as.data.frame(normalized_betas_sesame)  #putting these betas into a data frame format
rownames(raw)<-raw$CGid #naming the rownames by the metadata
rm(normalized_betas_sesame) #unmount to increase code run speed
raw.t<-t(raw[,-1]) #Transformation so that basenames appear as rows, and CpG sites as columns
as_tibble(raw.t) #convert to tibble
```

Selecting only the Eptesicus samples for age estimations.

```{r}
tojoin <- read.csv("BaseRage.csv") #file has basename and age of each EF samples 

known <- tojoin$Basename #this pulls out the basenames of interest

epfknowna <- raw.t[known, ] #faster way to pull out bats of interest

knowna<- epfknowna %>%    #pipe in bats of interest 
  as_tibble(., rownames="Basename")%>%   #continue to use tibble and assign rownames
  gather(., key= "site", value = "m", -Basename) %>%   #ordered by site
  mutate_at(vars(site), as.factor)%>%   #changing site names as factor
 mutate_at(vars(m), as.numeric) %>%   #assigning beta values as numeric values
 left_join(tojoin, .) #this joins age, site, basename, beta value of known Eptesicus fuscus

which(is.na(knowna), arr.ind=TRUE) # to see if any NAs are present
saveRDS(knowna, file = "joined.RDS") #saving as RDS for future analyses

```

Generating residuals for each individual by CpG site.

```{r}
library(dplyr)
library(tidyr)
library(modelr)
library(tidyverse)


#construct model to be used
fit_mod<- function(df) {
  lm(m ~ RevAge, data = df)
}

revageadj<- knowna %>%   
  group_by(site) %>%
  nest() %>% #organizes by CpG site
  mutate(model = map(data, fit_mod), #applying the knownt to the model
         resids= map2(data, model, modelr::add_residuals)) # setting up the linear model to pull residuals

batty<- revageadj %>% unnest(., resids) #unnesting to access the residuals

speedbat<- batty %>% select(-data, -model) #removing other parameters for faster load time

```


Below is code examining the linear regression that produced residuals at each site. This code generates an intercept, coefficient, pvalue, test statistic for each site.
```{r}
modelr.tidy<- revageadj %>%
  mutate(tidied = map(model, broom::tidy))

rat.tidy <- modelr.tidy %>%
  unnest(tidied) %>%
  select(site, term, estimate, std.error, statistic, p.value) %>%
  mutate_at(vars(term), factor) %>%
    mutate(., term = fct_recode(term, "Intercept" = "(Intercept)")) %>%
    pivot_wider(.,names_from = term, values_from = c(estimate, std.error, statistic, p.value))#estimate_RevAge is the model coefficient for age

write.csv(rat.tidy, "residualcoef.csv") 
```

Paired analysis:
Now paired residuals are filtered out and the band information is then added to the residuals of interest.

```{r}
library(readxl)
library(rstatix)
library(tidyverse)


names <- read.csv("Band_Basename.csv",na.strings = c("", "NA")) %>% na.omit 
keeper.resid<- speedbat %>% filter(Basename %in% names$Basename) #Filter applied to include only paired individuals

hib<-full_join(keeper.resid, names) #joining residuals to basename

#The following bats were sampled three times. In order to examine winter to summer, the middle sample was duplicated for these bats using the code below

Grey49d<- hib %>% filter(Band =="Grey49" & Time %in% c("S", "W21")) %>% #selecting season to duplicate
  ungroup(.) %>%
  mutate(Band =  "Grey49.1",
         Direction = "SW") %>% #adding the direction here
  mutate(Basename = recode(Basename,
                           `204529320118_R02C01` = "204529320118_R02C01.1",
                           `205483080041_R05C01` = "205483080041_R05C01.1")) #recoding the basename to indicate this bat was sampled three times

Red257d<- hib %>% filter(Band =="Red257" & Time %in% c("S", "W21")) %>%
  ungroup(.) %>%
  mutate(Band =  "Red257.1",
         Direction = "SW") %>%
  mutate(Basename = recode(Basename,
                           `204529320125_R06C01` = "204529320125_R06C01.1",
                           `205483080041_R04C01` = "205483080041_R04C01.1"))

hib.resids<-  hib %>%
  filter(!(Band %in% c("Grey49", "Red257") & Time =="W21")) %>%
  bind_rows(., Grey49d, Red257d) %>%
  mutate(., Direction = ifelse(Band %in% c("Grey49", "Red257"), "WS", Direction)) #joining these samples to the residual dataset

saveRDS(hib.resids, file = "hib.resid.RDS") #save as updated residuals
write.csv(hib.resids, "pairedresidual.csv") #contains all residuals in a csv file
```

Modeling season and remove bats that had died or moved into room temperature before samples were taken.
```{r}

sub.hib <- hib.resids%>%  
 filter(!Band %in% c( "Red257.1", "Blue42", "Blue58", "Blue52")) #Deceased bats or those that were moved into room temperature


par2 <- sub.hib %>%
  ungroup(.)%>%
  select(site, m, resid, Season, Band) %>% #selecting specific variables to speed up analyses
  mutate(Season = recode(Season, W = "first", S = "second")) %>% #this ensures that the analysis is run from winter to summer
  nest_by(site) %>% #Runs the analysis by site 
  mutate(model = list(nlme::lme(resid ~ Season, random = ~1|Band, data=data)))

tidy2<-par2 %>% select(site, model) %>% as_tibble(.) %>% #used to pull out relevant outputs from the linear model
  mutate(new.tidy = lapply(model, function(x) broom.mixed::tidy(x, effects="fixed"))) %>%
  unnest(new.tidy)%>%
  select(-model)


new2<-tidy2 %>%
  mutate(term= as_factor(term)) %>%
  mutate(., term = fct_recode(term, "Intercept" = "(Intercept)")) %>% #Recoding intercept variable so it is legible
  pivot_wider(.,names_from = c(term), values_from = c(estimate, std.error, df, statistic, p.value)) #Selects the variables for the final csv output file

new2$p.adjBY_season <- p.adjust(new2$p.value_Seasonsecond, method = "BY") #Running BY adjustment

write.csv(new2, "pairsseasonout.csv")
```

--------
Below, is example code used for plotting the residuals from winter and summer samples.
```{r}
library(ggpubr)
library(dplyr)
library(ggplot2)

pair.plot1 <- sub.hib %>% 
   as_tibble(.) %>%
  select(site, m, resid, Season, Band, Direction) %>%
  mutate(Season = factor(Season, levels=c("S","W")))

#cg10775708, Distal intergenic, MAZ, winter up
wup4 <- pair.plot1 %>% filter(site=="cg10775708")

ggpaired(wup4, x = "Season", y = "resid", id="Band", 
         fill = "Season", line.color = "gray", line.size = 0.2,
         palette = c("orange", "blue"),
         title = "cg10775708",
         xlab="Season", ylab="Residual Methylation",
legend = "none", ylim = c(-0.2, 0.13))
```